%%% How The RBM->UBM transition arises naturally when implementing Gates %%%


\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

\begin{document}
%
%
\section{Unitary Operations and RBMs}
%
The state of an $N$-spin system is described by a wavefunction $\Psi(s)$ with
$s\in\{0,1\}^N$. $\Psi(s)$ can be encoded in a so-called Boltzmann Machine, a
simple two-layer ANN. It consists of $N_v$ visible and $N_h$ hidden nodes and
is specified by parameters $a\in\mathbb{R}^{N_v}, b\in\mathbb{R}^{N_h}$ and
$w\in\mathbb{R}^{N_h\times N_v}$. The energy function
\begin{equation}
    E(v,h) = v^Ta + h^Tb + h^Twv
\end{equation}
allows one to express probability distributions defined on $\mathbb{R}^{N_v}$
according to
\begin{equation}
    P(v) = \frac{\sum_{h}e^{-E(v,h)}}{\sum_v \sum_h e^{-E(v,h)}}.
\end{equation}
Similarly, if we allow complex $a\in\mathbb{C}^{N_v}, b\in\mathbb{C}^{N_h}$ and
$w\in\mathbb{C}^{N_h\times N_v}$ we can map each spin configuration
$s\in\{0,1\}^{N_S}$ of system with $N_S$ spins to a complex amplitude. This is
then the wavefunction
\begin{equation}
    \Psi(s) = \frac{\sum_{h}e^{-E(s,h)}}{\sum_s \sum_h e^{-E(s,h)}}.
    \label{eq:RBM-psi}
\end{equation}
Consider now a 1-body unitary operator $O\in U(2)$. It is completely specified
by its four matrix elements $O_{ss'}$ where $s,s'\in\{0,1\}$. Equivalently, we
can express this Operator in terms of an exponential function:
\begin{equation}
    O(s, s') = A \exp{(\alpha s + \beta s' +  \omega ss')}.
\end{equation}
To do this we simply associate
\begin{align*}
    A &= O_{00} \\
    \alpha &= \ln{\left(\frac{O_{10}}{A}\right)} \\
    \beta &= \ln{\left(\frac{O_{01}}{A}\right)} \\
    \omega &= \ln{\left(\frac{O_{11}}{A}\right)} - (\alpha + \beta).
\end{align*}
Clearly the function $O(s,s')$ defined this way behaves like the operator. The
wavefunction $\Psi'(s)$ after $O$ has acted upon the initial state $Psi(s)$ is
given by
\begin{align}
    \Psi'(s) &= \sum_{s'} O_{ss'} \Psi(s_1,\hdots, s', \hdots, s_N) \nonumber \\
             &= \sum_{s'} \exp{(\alpha s + \beta s' +  \omega ss')} \Psi(s_1,
                \hdots, s', \hdots, s_N)
\end{align}
Expressing the RHS wave-function now in terms of our RBM (cf. \ref{eq:RBM-psi}),
the sum over spin $s'$ can be associated with another hidden node.
%
%
%
\section{Unitary Operations and POVM Distributions}
Suppose now the RBM encodes some probability distribution $P(\vec{a})$ of an
informationally complete POVM measurement. This distribution then uniquely
specifies some quantum state. Let $O_{\vec{a},\vec{b}}$ now denote a single-body
unitary operator acting on POVM distribution vectors. Is it still possible to
rewrite this operator as an exponential function and describe the effect of the
unitary operation as the insertion of an additional node into the RBM? \par
Let us first consider the operator. The principal goal is to write it in
exponential form such that it can be absorbed into the energy functional of a
slightly modified architecture. There exist several ways of implementing POVMs
outcoms in an RBM. For reasons which will become apparent later we will choose
binary encoding. This means each outcome is encoded by pairs of two nodes in the
network. Binary nodes then lead to the configurations in \ref{tab:node_conf}
%
\begin{figure}[h!]
  \centering
  \begin{tabular}{c|c}
      POVM outcome & node configuration \\
      \hline
      1 & (0,0) \\
      2 & (0,1) \\
      3 & (1,0) \\
      4 & (1,1)
  \end{tabular}
  \caption{4-POVM outcomes implemented by binary nodes.}
  \label{tab:node_conf}
\end{figure}
%
A single-qubit gate in the POVM formalism corresponds to a complex
($4\times4$)-matrix. Its 16 entries need to be represented in their entirety in
the exponential representation of this operator. Rows in this matrix can be
associated with a fixed ''target outcome'' while entries in the same column
correspond to the same ''initial outcome''. Let us denote the two nodes encoding
the target outcome by $(v_1, v_2)$ and the nodes encoding the initial outcome
by $(h_1, h_2)$. The entry $O_{ij}$ corresponds to the transition element for
$j\rightarrow i$, where $i,j\in\{1,\hdots,4\}$ represent POVM measurements.
These states can also be described in terms of the nodes that encode them. Then
we have transitions $(h^j_1,h^j_2)\rightarrow(v^i_1,v^i_2)$. Each transition
(to which one specific matrix element is associated) can be represented by a
unique 4-tuple $(h^j_1,h^j_2, v^i_1,v^i_2)$ where each entry takes binary
values.  \par
Consider now products of these nodes. All possible products are given by
\begin{equation}
    \{ h_1^i h_2^j v_1^k v_2^l \: : \: (i,j,k,l)\in \{0,1\}^4 \}
\end{equation}
I will argue that the value of these variable-products can be used to identify
specific transitions. It is clear that $h_1h_2v_1v_2$ only evaluates to 1 if all
variables are equal to one. Hence this outcome can be used to identify the
($4\rightarrow4$) transition. We hence include this product in the exponent and
multiply it with a coefficient $\ln{(O_{44})}$ including some correction factors
which we will discuss later. The energy functional we choose is
\begin{equation}
    E(v_1,v_2,h_1,h_2) = \sum_{i=0,1}\sum_{j=0,1}\sum_{k=0,1}\sum_{l=0,1}
                         c_{ijkl}h_1^i h_2^j v_1^k v_2^l
\end{equation}
We recognize that the transition $1\rightarrow1$ encoded by binary variable
products in the energy functional is just $c_{0000}$ and always evaluates to 1,
irrespective of the specific input $(v_1,v_2,h_1,h_2)$. We thus write
$c_{0000}=\ln(O_{11})$. For input $(0,0,0,0)$ all other terms don't contribute
since at least one factor is always zero. When looking at configurations that
involve some nodes not equal to zero, we have to make the following consideration:
Say we look at the transition $4\rightarrow1$. This corresponds to $v_1=0,v_2=0
h_1=1,h_2=2$. All products unequal to zero are then $h1, h2, h1h2$. We cannot
simply identify $c_{0011} = O_{14}$, but instead have to account for the fact
that we have ''unwanted'' activation products ($h1,h2$ corresponding to the
transitions $2\rightarrow1, 3\rightarrow1$). Thus we exclude these by an
appropriate coefficient choice: $c_{0011} = O_{14} - c_{0010} - c_{0001}$. This
procedure can be applied to every $c_{ijkl}$ and we end up with a unique relation
$c_{ijkl} = f_{ijkl}(O)$. \par
Since we showed that the POVM-Operator can be represented as an exponential
function, the occuring sum over ''prior'' outcomes $b_i$ can be -- together with
the operator in exponent form -- associated with additional hidden nodes added
to the network. For our one-body operation, this leads to already two (rather
than 1 for the case of spin operators) additional hidden nodes. These interact
with other hidden nodes and hence we have an Unrestricted Boltzmann machine.
This makes information extraction complicated, since no analytical expression
for the probability encoded by the UBM can be found anymore due to the
interactions of the hidden nodes. What can be done however is to transform the
unrestricted Boltzmann machine into a Deep Boltzmann Machine by the addition of
a second hidden layer that includes the two nodes added by the one-body
operation. What were formally interactions within the hidden layer are now
interaction between the first and second hidden layer. The two additional nodes
however also interact with the visible units. As we do not allow direct
interaction between the visible and second hidden layer in the DBM, additional
nodes can be inserted in the first hidden layer that mediate the interaction
between the visible and second hidden layer. \par
Then information from the DBM could be extracted as follows:
\begin{enumerate}
  \item Consider hidden layer 1 and hidden layer 2 as a simple RBM. The sum over
  the units from hidden layer 2 can hence be performed analytically.
  \item Perform MC-sampling for the hidden nodes from hidden layer 1.
\end{enumerate}
The question that arises now is whether the new operator expression still allows
such a procedure. The links in the graph structure that is the Boltzmann machine
are determined by the form of the energy functional. Interactions can be
classified as follows:
\begin{itemize}
  \item One-point functions: For instance $h_1, v_2, \hdots$. Their coefficients
  correspond to biases associated with the corresponding nodes.
  \item Two-point functions: Here three cases need to be distringuished
  \begin{itemize}
    \item Visible-hidden: Such as $v_1h_2$. These are allowed and can be realized
    by weights alone while still agreeing with the RBM structure.
    \item Visible-Visible: Such as $v_1v_2$. These interactions appear
    irreconcilable with the RBM structure at first but actually can be realized
    if we simply introduce another hidden node that mediates their interaction.
    \item Hidden-Hidden: Such as $h_1h_2$. These interactions are problematic
    since they forbid efficient sampling from the NN. Just as the interactions
    in the visible layer, these can be impemented by a new node in another,
    third layer. This circumvents intralayer connections but does not make
    sampling any easier.
  \end{itemize}
  \item Three point functions: Such as $v_1v_2h_1$. Not sure yet if they can be
  rewritten in terms of simpler functions.
  \item Four-point function: This is just $v_1v_2h_1h_2$ and is very complicated
  to implement.
\end{itemize}
While we have shown that it is generally possibe to rewrite single-qubit gates
in terms of POVM distributions, they lead to energy functionals that are too
complicated to be implemented efficiently. Hence we conclude that it is not
worth pursuing this approach.
\end{document}
